---
output: github_document
---

# ALE: Accumulated Local Effects for Interpretable Machine Learning

`ALE` is lighweight R package designed to compute and visualize [Accumulated Local Effects (ALE)](https://arxiv.org/pdf/1612.08468), providing interpretability for machine learning models. 

## Installation

`ALE` package is not yet on the CRAN, so you can install it directly from github: 

```{r}
#install.packages("devtools")
devtools::install_local("ielbadisy/ALE")
```


## Usage

### Loading the Package

```{r}
library(ALE)
```

### Computing ALE

Use the `compute_ale()` function to calculate ALE for a feature in your model. Here’s a quick example:

```{r}
library(ALE)

# Simulate some data
set.seed(123)
data <- data.frame(x1 = runif(100), x2 = runif(100))
data$y <- 4 * data$x1 + 3 * data$x2 + rnorm(100)

# Fit a model
model <- lm(y ~ x1 + x2, data = data)

# Define the prediction function explicitly
pred_fun <- stats::predict

# Compute ALE for x1, specifying the prediction function explicitly
ale_x1 <- compute_ale(model, data, "x1", grid_size = 10, pred_fun = pred_fun)
ale_x1
```

### Plotting ALE

Use the `plot_ale()` function to visualize the ALE values for a feature:

```{r}
# Plot ALE for x1
plot_ale(ale_x1, feature_name = "x1", title_suffix = "Custom Model")
```

### Comparative Analysis with ALEPlot Package

If you wish to compare ALE values computed with this package to those from the `ALEPlot` package, you can do so with the following steps:

```{r}
# define a custom prediction function with the expected argument names
yhat <- function(X.model, newdata) {
    predict(X.model, newdata = newdata)
}

# now call ALEPlot with the corrected prediction function
ALEPlot_data <- ALEPlot::ALEPlot(data[, c("x1", "x2")], model, pred.fun = yhat, J = 1, K = 10)
```

## Documentation

Full documentation for `ALE` functions, including additional examples and parameters, can be found in the package help files:

```{r}
?compute_ale
?plot_ale
```

Or view the package vignette:

```{r}
vignette("example", package = "ALE")
```

## Contributing

Contributions are welcome! Please follow these steps:

1. Fork the repository.
2. Create a new branch for your feature or bugfix.
3. Submit a pull request with a clear description of your changes.

## License

This package is licensed under the [License: GPL v3](https://img.shields.io/badge/license-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0).
