---
title: "Example: Analyzing Accumulated Local Effects (ALE)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to compute and visualize Accumulated Local Effects (ALE) using the ALE package. ALE plots help to interpret the effects of predictors in a model, showing how features influence the prediction on average, locally across their range.

## Setup

Load necessary libraries and prepare the environment:

```{r setup}
library(ALE)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(ggpubr)
library(nnet)  # for neural network modeling
library(ALEPlot)  # alternative ALE plotting

# Create synthetic data for demonstration
set.seed(42)
n <- 5000
x1 <- runif(n, min = 0, max = 1)
x2 <- runif(n, min = 0, max = 1)
x3 <- runif(n, min = 0, max = 1)
x4 <- runif(n, min = 0, max = 1)
y <- 4*x1 + 3.87*x2^2 + 2.97*exp(-5 + 10*x3) / (1 + exp(-5 + 10*x3)) + 
     13.86*(x1 - 0.5)*(x2 - 0.5) + rnorm(n, 0, 1)
DAT <- data.frame(y, x1, x2, x3, x4)

# Build a neural network model
nnet.DAT <- nnet(y ~ ., data = DAT, linout = TRUE, skip = FALSE, size = 8,
                 decay = 0.1, maxit = 1000, trace = FALSE)
```

## Predictive Function

```{r}
# Define a predictive function for the neural network model
yhat <- function(X.model, newdata) as.numeric(predict(X.model, newdata, type = "raw"))
```

## Compute ALE for Each Predictor

```{r}
# Compute ALE using custom function for each predictor
ale_custom_x1 <- compute_ale(nnet.DAT, DAT, "x1", grid_size = 100)
ale_custom_x2 <- compute_ale(nnet.DAT, DAT, "x2", grid_size = 100)
ale_custom_x3 <- compute_ale(nnet.DAT, DAT, "x3", grid_size = 100)
ale_custom_x4 <- compute_ale(nnet.DAT, DAT, "x4", grid_size = 100)
```

## Plot ALE Results

Generate plots for each feature's ALE:

```{r plot-ale}
plot_custom_x1 <- plot_ale(ale_custom_x1, "x1", "Custom")
plot_custom_x2 <- plot_ale(ale_custom_x2, "x2", "Custom")
plot_custom_x3 <- plot_ale(ale_custom_x3, "x3", "Custom")
plot_custom_x4 <- plot_ale(ale_custom_x4, "x4", "Custom")

# Use ALEPlot package for comparison
ALE_1 <- ALEPlot(DAT[, 2:5], nnet.DAT, pred.fun = yhat, J = 1, K = 500, NA.plot = TRUE)
ALE_2 <- ALEPlot(DAT[, 2:5], nnet.DAT, pred.fun = yhat, J = 2, K = 500, NA.plot = TRUE)
ALE_3 <- ALEPlot(DAT[, 2:5], nnet.DAT, pred.fun = yhat, J = 3, K = 500, NA.plot = TRUE)
ALE_4 <- ALEPlot(DAT[, 2:5], nnet.DAT, pred.fun = yhat, J = 4, K = 500, NA.plot = TRUE)
```

## Comparative Visualization

Arrange the custom and ALEPlot results side by side for comparison:

```{r arrange-plots}
# Convert ALEPlot results to data frames for consistency in plotting
convert_ale_to_df <- function(ALE, feature) {
  data.frame(feature = ALE$x.values, ALE = ALE$f.values)
}
aleplot_x1 <- convert_ale_to_df(ALE_1, "x1")
aleplot_x2 <- convert_ale_to_df(ALE_2, "x2")
aleplot_x3 <- convert_ale_to_df(ALE_3, "x3")
aleplot_x4 <- convert_ale_to_df(ALE_4, "x4")

# Plot ALE results from ALEPlot package
plot_aleplot_x1 <- plot_ale(aleplot_x1, "x1", "ALEPlot")
plot_aleplot_x2 <- plot_ale(aleplot_x2, "x2", "ALEPlot")
plot_aleplot_x3 <- plot_ale(aleplot_x3, "x3", "ALEPlot")
plot_aleplot_x4 <- plot_ale(aleplot_x4, "x4", "ALEPlot")

# Combine plots for each predictor
combined_plot_x1 <- ggarrange(plot_custom_x1 + ggtitle("Custom ALE for x1"), plot_aleplot_x1 + ggtitle("ALEPlot for x1"), ncol = 2)
combined_plot_x2 <- ggarrange(plot_custom_x2 + ggtitle("Custom ALE for x2"), plot_aleplot_x2 + ggtitle("ALEPlot for x2"), ncol = 2)
combined_plot_x3 <- ggarrange(plot_custom_x3 + ggtitle("Custom ALE for x3"), plot_aleplot_x3 + ggtitle("ALEPlot for x3"), ncol = 2)
combined_plot_x4 <- ggarrange(plot_custom_x4 + ggtitle("Custom ALE for x4"), plot_aleplot_x4 + ggtitle("ALEPlot for x4"), ncol = 2)

# Display combined plots
grid.arrange(combined_plot_x1, combined_plot_x2, combined_plot_x3, combined_plot_x4, nrow = 4)
```
## All in one plot

Fist we extract and prepare the data for plotting: 
```{r}
aleplot_x1 <- convert_ale_to_df(ALE_1, "x1")
aleplot_x2 <- convert_ale_to_df(ALE_2, "x2")
aleplot_x3 <- convert_ale_to_df(ALE_3, "x3")
aleplot_x4 <- convert_ale_to_df(ALE_4, "x4")

ale_custom_x1$Method <- "Custom"
ale_custom_x2$Method <- "Custom"
ale_custom_x3$Method <- "Custom"
ale_custom_x4$Method <- "Custom"

aleplot_x1$Method <- "ALEPlot"
aleplot_x2$Method <- "ALEPlot"
aleplot_x3$Method <- "ALEPlot"
aleplot_x4$Method <- "ALEPlot"

# combine the data
combined_x1 <- rbind(ale_custom_x1, aleplot_x1)
combined_x2 <- rbind(ale_custom_x2, aleplot_x2)
combined_x3 <- rbind(ale_custom_x3, aleplot_x3)
combined_x4 <- rbind(ale_custom_x4, aleplot_x4)


```

Then we can plot: 
```{r}
library(ggplot2)
library(gridExtra) 

# function to plot ALE for each feature
plot_ale_comparison <- function(data, feature_name) {
  ggplot(data, aes(x = feature, y = ALE, color = Method)) +
    geom_line() +
    labs(title = paste("ALE Comparison for", feature_name),
         x = feature_name, y = "Accumulated Local Effect (ALE)") +
    theme_minimal() +
    scale_color_manual(values = c("Custom" = "blue", "ALEPlot" = "red")) +
    theme(legend.position = "bottom")
}

# create plots for each feature
plot_x1 <- plot_ale_comparison(combined_x1, "x1")
plot_x2 <- plot_ale_comparison(combined_x2, "x2")
plot_x3 <- plot_ale_comparison(combined_x3, "x3")
plot_x4 <- plot_ale_comparison(combined_x4, "x4")

# arrange the plots in a grid
grid.arrange(plot_x1, plot_x2, plot_x3, plot_x4, ncol = 2)
```
